### 2024-04-20 MSRI-16-1989-79.Rmd
more pages

::: {.rmenv #chunk-1.5 title="1.5 Remarks" latex="{1.5 Remarks}"}
—

i. If $M_\DR$ is given, then the data of $M_\BB$, $F_\infty$, and $\comp_{\DR,\BB}$ satisfying [(AM2)](#AM2) is equivalent to that of a new rational structure $M_\BB\subset M_\DR\otimes\CC$ that is stable under complex conjugation $c_\DR$ (set $F_\infty=c_\DR|M_\BB$).
  By [(M6)](#M6), the filtration $W$ of $M_\DR$ must remain rational for this new rational structure.

ii. The data of $M_\AAf$, $\comp_{\AAf,\BB}$, and the Galois action, all together, are equivalent to the data of a $\QQ_\ell$-vector space $M_\ell$ for all $\ell$, along with a Galois action on $M_\ell$ and comparison isomorphisms $\comp_{\ell,\BB}\colon M_\BB\otimes\QQ\simto M_\ell$.
  We have to assume the existence of a lattice $L\subset M_\BB$ such that the $\comp_{\ell,\BB}(L\otimes\ZZ_\ell)$ are Galois stable.
  We define $M_\AAf$ from the $M_\ell$ as the restricted product of the $M_\ell$ with respect to the $\comp_{\ell,\BB}(L\otimes\ZZ_\ell)$ for an arbitrary lattice $L$: this restricted product is Galois stable, and the $\comp_{\ell,\BB}$ induce $\comp_{\AAf,\BB}$.

  The data of $M_\ell$, $\comp_{\ell,\BB}$, and the Galois action (resp. $M_\AAf$, $\comp_{\AAf,\BB}$, and the action), all together, are also equivalent to the data of a Galois action on $M_\BB\otimes\QQ_\ell$ (resp. $M_\BB\otimes\AAf$).
  By [(M6)](#M6) and [(M9)](#M9), the filtration of $M_\BB\otimes\QQ_\ell$ (resp. $M_\BB\otimes\AAf$) induced by $W$ must be stable under $\Gal(\bar{\QQ}/\QQ)$.

iii. If $M_\DR$ is given, then the data of $M_{\crisp}$, along with its crystalline Frobenius and $\comp_{\crisp,\DR}$, is equivalent to the data of an automorphism $\phi_p$ of $M_\DR\otimes\QQ_p$.
  By [(M6)](#M6) and [(M10)](#M10), the filtration of $M_\DR\otimes\QQ_p$ induced by $W$ must be stable under $\phi_p$.

We will often tacitly use these remarks to describe a system [(M1)--(M10)](#chunk-1.4).
:::

::: {.rmenv #chunk-1.6 title="1.6" latex="{1.6}"}
A scheme $X$ of finite type over $\QQ$ should define, for each $i$, a motive $M\coloneqq \HH_\mot^i(X)$.
\oldpage{95}In this section, we will partially describe the system [(M1)--(M10)](#chunk-1.4) of realisations of $M$ in the case where $X$ is separated and smooth over $\QQ$.

We have $M_\BB=\HH^i(X(\CC),\QQ)$, and $F_\infty$ is induced by the complex conjugation of $X(\CC)$;
$M_\DR=\HH_\DR^i(X)\coloneqq\bHH^i(X,\Omega_X^\bullet)$, and the Hodge filtration that that of the mixed Hodge theory [@D2, Section 3.2];
$M_\ell=\HH^i(X\otimes\bar{\QQ},\QQ_\ell)$ is the $\ell$-adic étale cohomology of the scheme over $\bar{\QQ}$ induced from $X$ by extension of scalars, and the action of $\Gal(\bar{\QQ},\QQ)$ is given by structure transport.
_Notation:_ $X\otimes\bar{\QQ}$, cf. [(0.4)](#chunk-0.4).

Suppose that $X$ is smooth and proper, and let $S$ be a finite set of prime numbers such that $X$ is the general fibre of $X^\sim$, which is smooth and proper over $\Spec(\ZZ)\setminus S$.
For $p\not\in S$, $M_\crisp$ is the crystalline cohomology of the reduction $X^\sim\otimes\FF_p$ of $X$ modulo $p$, tensored over $\ZZ_p$ with $\QQ_p$.
The crystalline Frobenius $\phi_p$ is induced by the inverse image morphism of the Frobenius $\Fr\colon X^\sim\otimes\FF_p\to X^\sim\otimes\FF_p$.

More generally, suppose that we have some smooth and proper $\overline{X}$ over $\Spec(\ZZ)\setminus S$, as well as a relative normal crossing divisor $D$;
let $X$ be the general fibre of $\overline{X}\setminus D$.
Then the realisation $M_\crisp$ is defined for $p\not\in S$;
its most natural definition is given by the generalisation of the crystalline theory, considered by Faltings in [@Fa, IV], to the "logarithmic poles" case.

The comparison isomorphism $\comp_{\DR,\BB}$ is (1.1.3), and the comparison isomorphism $\comp_{\ell,\BB}$ is (1.1.4).

In the smooth and proper case, the comparison isomorphism $\comp_{\crisp,\DR}$ comes from §7.26 of [P. Berthelot and A. Ogus, _Notes on crystalline cohomology_, Princeton University Press and Tokyo University Press, 1978].
For the general case, see [@Fa, IV].
Finally, the weight filtration $W$ is that of the mixed Hodge theory from [@D2, Section 3.2].
See also [@D5].
:::

::: {.rmenv #chunk-1.7 title="1.7" latex="{1.7}"}
An additional data that we have on the cohomology $M\coloneqq\HH_\mot^i(X)$ when $X$ is smooth over $\QQ$ is that of a comparison isomorphism, for almost all $p$, in the sense of Fontaine--Messing (cf. [@FM; @Fa]), relating $M_p$, endowed with the action of a decomposition group of $p$, to $M_\DR\otimes\QQ_p$, endowed with its Hodge filtration and its crystalline Frobenius.

For all $p$, we should also have a "crystalline" structure of the following type.

- **Semi-stable case.**
  Let $T_p$ be the Zariski tangent space of $\Spec(\ZZ_p)$ at its closed point.
  \oldpage{96}We complete it to a projective line $\overline{T}_p$ over $\FF_p$, and we can lift $(\overline{T}_p,0,\infty)$ to a projective line endowed with two marked points over $\ZZ_p$: $(\overline{T}_p^\sim,0,\infty)$.
  We want an $F$-isocrystal with logarithmic poles on $(\overline{T}_p,0,\infty)$ (cf. [@Fa]).
  Such an object induces, on $\overline{T}_p^\sim\otimes\QQ_p$, a module with connection $\mathcal{V}$ with logarithmic poles at $0$ and at $\infty$, and we want for the residue of the connection at $0$ and at $\infty$ to be nilpotent.
  If $\varphi$ is a section of $\overline{T}_p^\sim$, over $\Spec(\ZZ_p)$, with derivative equal to $1$ at the closed point, then $\varphi^*\mathcal{V}$ is independent of the choice of $\varphi$, and $\comp_{\DR,\crisp}$ should then be identified with the de Rham realisation $\otimes\QQ_p$.

- **General case.**
  The data of the previous type, over a large-enough finite Galois extension $E$ of $\QQ_p$ that is $\Gal(E/\QQ_p)$-equivariant.

A Fontaine--Messing comparison isomorphism should again link this object and $M_p$ endowed with the action of a decomposition group of $p$.
:::

::: {.rmenv #chunk-1.8 title="1.8 Variant" latex="{1.8 Variant}"}
We should also have [(M1)--(M10)](#chunk-1.4) for $M\coloneqq\HH_\mot^i(X)$, where $X$ is not necessarily smooth.
The crystalline data pose a problem.

We would also like to have [(M1)--(M10)](#chunk-1.4) for cohomology with proper support.
:::

::: {.rmenv #chunk-1.9 title="1.9 Definition" latex="{1.9 Definition}"}
A _realisation system_ is a system [(M1)--(M10)](#chunk-1.4) that satisfies [(A1)--(A5)](#chunk-1.4).
:::

We understand [(M4)](#M4), $\comp_{\crisp,\DR}$, and [(M10)](#M10) as a germ --- in the filter of complements of finite sets of prime numbers --- of systems of automorphisms $\phi_p$ of the $M_\DR\otimes\QQ_p$

::: {.itenv #chunk-1.10 title="Proposition 1.10" latex="{Proposition 1.10}"}
Realisation systems form a Tannakian category.
:::

::: {.proof}
As in [@J], the key point is that a morphism of mixed Hodge $\QQ$-structures is strictly compatible with the filtrations $W$ and $F$, and that its kernel and cokernel are mixed Hodge $\QQ$-structures [@D2, Theorem 2.3.5].
We thus deduce that every morphism of realisation systems is strictly compatible with $W$ and $F$.
(For $W$, we can instead use [(AM4)](#AM4).)

So it is clear that the kernels and cokernels again form realisation systems, and that a bijective morphism is an isomorphism.
We have direct sums, and so the category is abelian.

We have an obvious tensor product, which is associative and commutative, and a way of taking duals.
We also have a fibre functor, or, indeed, two: ${}_\BB$ and ${}_\DR$, with values in $\QQ$-vector spaces.
The category of realisation systems is thus Tannakian and **TO-DO**: the fibre functor ${}_\BB$ (resp. ${}_\DR$) identifies it with the category of representations of the group scheme $G_\BB$ (resp. $G_\DR$) of its automorphisms (cf. [@Sa] or [@DM, Theorem 2.11]).
:::

\oldpage{97}We hope that the realisation functors define a fully faithful functor from the category of motives over $\QQ$ to the category of realisation systems.
If this were not the case, then the philosophy of motives would lose much of its interest.
This leads to the following provisional "definition":

::: {.rmenv #chunk-1.11 title="1.11 Definition" latex="{1.11 Definition}"}
The _category of motives_ over $\QQ$ is the subcategory of the category of realisation systems [(1.9)](#chunk-1.9) generated (under $\oplus$, $\otimes$, dual, and sub-quotient) by the category of systems of geometric origin.
:::

This definition is not really a definition, since "of geometric origin" has not been defined.
Worse still, I do not have any definition to propose that I can confidently say is good.

::: {.rmenv #chunk-1.12 title="1.12 Remark" latex="{1.12 Remark}"}
We would hope that the realisations of motives have properties not included in the definition [(1.9)](#chunk-1.9) of realisation systems.
Some reasons for not including them:

a. we do not know how to verify them in practice;
b. we no longer know how to prove [(1.10)](#chunk-1.10) if we do include them.

Thus.

i. We would like that, for almost all $p$, the Frobenius characteristic polynomial $\det(1-F_pt,M_\ell)$ at $p$ have rational coefficients that are independent of $\ell\neq p$.
  It should also agree with $\det(1-\phi_pt,M_\crisp)$.
  We do not know how to verify this for $\HH_\mot^i(X)$ (with $X$ smooth), nor for a direct factor of $\HH_\mot^i(X)$ (with $X$ an abelian variety), and reason (b) above also applies.

ii. We would like to complete [(AM4)](#AM4) by a condition for all $p$, cf. 1.8.5 in [P. Deligne, "La conjecture de Weil II", _Publ. Math. IHÉS_ **52** (1980), 137--252].
  Reasons (a) and (b) above also apply.

iii. The Hodge structure $\Gr_n^W(M_\BB)$ should be polarisable.
  More precisely, there should exist, for all $n$, a morphism of realisation systems
  $$
    \Gr_n^W(M)\otimes\Gr_n^W(M) \to \QQ(-n)
  $$
  (see [(2.1)](#chunk-2.1) for the definition of $\QQ(-n)$) that induces a polarisation of the weight-$n$ Hodge structure $\Gr_n^W(M_\BB)=\Gr_n^W(M_\BB)$.
  Here, neither reason (a) nor reason (b) apply.
:::

::: {.rmenv #chunk-1.13 title="1.13 Remark" latex="{1.13 Remark}"}
\oldpage{98}The treatment of crystalline structures is not satisfying.
In the definition of realisation systems, I have not included the data  given in [(1.7)](#chunk-1.7), for the want of verifying their existence in the case of Lie algebras of $\pi_1$ that interest us.
I have nevertheless included the data of $\phi_p$, despite its appearance as a bizarre addition, because the calculations in [§19](#section-19) give an interesting result.
:::

@@

A definition essentially equivalent to [(1.9)](#chunk-1.9) is then the following: a _realisation system_ is a system [(M'1)](#Mp1) to [(M'6)](#Mp6), and [(M'8)](#Mp8), satisfying axioms [(AM1)](#AM1), [(AM2)](#AM2), and [(AM4)](#AM4) (suitably modified in the evident way).

@@

::: {.rmenv #chunk-1.15 title="1.15 Variant" latex="{1.15 Variant}"}
Let $P$ be a finite set of prime numbers.
The category of *smooth realisation systems on $\Spec(\ZZ)\setminus P$* is defined as in [(1.9)](#chunk-1.9), taking $P$ to be the exceptional set $S$ in [(AM3)](#AM3), and replacing "almost all $p$" in [(M4)](#M4) and [(AM4)](#AM4) by "all $p\not\in P$".
This treatment of crystalline structures is not satisfying, cf. [(1.7)](#chunk-1.7).
The category [(1.9)](#chunk-1.9) of realisation systems on $\Spec(\QQ)$ is the inductive limit of these categories for $P$ growing larger and larger.

@@

::: {.rmenv #chunk-2.1 title="2.1" latex="{2.1}"}
\oldpage{105}The *Tate motive* $\ZZ(1)$ is a motive over $\Spec(\ZZ)$ [(1.15)](#chunk-1.15) with integer coefficients [(1.23)](#chunk-1.23).
Here is its description as a realisation system, in the language of [(1.9)](#chunk-1.9) and [(1.14)](#chunk-1.14) (along with [(1.25)](#chunk-1.25)):

@@

::: {.rmenv #chunk-2.4 title="2.4" latex="{2.4}"}
I conjecture that the set of motives with integer coefficients of the form $T(X)$ for some $1$-motive $X$ is stable under extensions.
If $T'$ is a motive with integer coefficients, with $T'\otimes\QQ\simto T(X)\otimes\QQ$, then $T'$ is of the form $T(X')$ with $X'$ isogenous to $X$.
The conjecture is thus equivalent to the claim that the set of motives $T(X)\otimes\QQ$, for $1$-motives $X$, is stable under extensions.
The word "conjecture" is an abuse of terminology, since the statement itself is not precise.
What is conjectured is that every realisation system [(1.9)](#chunk-1.9) (or [(1.24)](#chunk-1.24), over $S$) that is an extension $T(X)$ by $T(Y)$ (for $1$-motives $X$ and $Y$), and "natural", "coming from geometry", is isomorphic to that defined by a $1$-motive $Z$ that is an extension of $X$ by $Y$.

@@

:::: {.rmenv #chunk-2.5 title="2.5" latex="{2.5}"}
Let $M$ be a realisation system [(1.9)](#chunk-1.9).
We spell out what a realisation system $E$ given by an extension of $\QQ(0)$ by $M$ is, when $M$ is of weight $\leqslant-1$ (i.e. $M=W_{-1}(M)$).

@@

::: {.trmenv #M5P title="(M5)P" latex="{(M5)P}"}
The comparison morphisms for $E$ can be identified with comparison morphisms $P_\BB\otimes\CC\simto P_\DR\otimes\CC$, $P_\BB\otimes\AAf\simto P_\AAf$, and $P_\DR\otimes\QQ_p\simto P_\crisp$ that are compatible with the analogous morphisms for $M$.
*Notation:* $\otimes$, see [(0.4)](#chunk-0.4).
:::

>>>>

::: {.rmenv #chunk-1.13 title="1.13 Remark" latex="{1.13 Remark}"}
\oldpage{98}
The treatment of crystalline structures is not satisfying.
In the definition of realisation systems, I have not included the data  given in [(1.7)](#chunk-1.7), for the want of verifying their existence in the case of Lie algebras of $\pi_1$ that interest us.
I have nevertheless included the data of $\phi_p$, despite its appearance as a bizarre addition, because the calculations in [§19](#section-19) give an interesting result.
:::

@@

A definition essentially equivalent to \ref{minus-three-points-1.9} is then the following: a _realisation system_ is a system [(M'1)](#Mp1) to [(M'6)](#Mp6), and [(M'8)](#Mp8), satisfying axioms [(AM1)](#AM1), [(AM2)](#AM2), and [(AM4)](#AM4) (suitably modified in the evident way).

@@

::: {.rmenv #chunk-1.15 title="1.15 Variant" latex="{1.15 Variant}"}
Let $P$ be a finite set of prime numbers.
The category of *smooth realisation systems on $\Spec(\ZZ)\setminus P$* is defined as in \ref{minus-three-points-1.9}, taking $P$ to be the exceptional set $S$ in [(AM3)](#AM3), and replacing "almost all $p$" in [(M4)](#M4) and [(AM4)](#AM4) by "all $p\not\in P$".
This treatment of crystalline structures is not satisfying, cf. [(1.7)](#chunk-1.7).
The category \ref{minus-three-points-1.9} of realisation systems on $\Spec(\QQ)$ is the inductive limit of these categories for $P$ growing larger and larger.

@@

::: {.rmenv #chunk-2.1 title="2.1" latex="{2.1}"}
\oldpage{105}The *Tate motive* $\ZZ(1)$ is a motive over $\Spec(\ZZ)$ [(1.15)](#chunk-1.15) with integer coefficients [(1.23)](#chunk-1.23).
Here is its description as a realisation system, in the language of \ref{minus-three-points-1.9} and [(1.14)](#chunk-1.14) (along with [(1.25)](#chunk-1.25)):

@@

::: {.rmenv #chunk-2.4 title="2.4" latex="{2.4}"}
I conjecture that the set of motives with integer coefficients of the form $T(X)$ for some $1$-motive $X$ is stable under extensions.
If $T'$ is a motive with integer coefficients, with $T'\otimes\QQ\simto T(X)\otimes\QQ$, then $T'$ is of the form $T(X')$ with $X'$ isogenous to $X$.
The conjecture is thus equivalent to the claim that the set of motives $T(X)\otimes\QQ$, for $1$-motives $X$, is stable under extensions.
The word "conjecture" is an abuse of terminology, since the statement itself is not precise.
What is conjectured is that every realisation system \ref{minus-three-points-1.9} (or [(1.24)](#chunk-1.24), over $S$) that is an extension $T(X)$ by $T(Y)$ (for $1$-motives $X$ and $Y$), and "natural", "coming from geometry", is isomorphic to that defined by a $1$-motive $Z$ that is an extension of $X$ by $Y$.

@@

:::: {.rmenv #chunk-2.5 title="2.5" latex="{2.5}"}
Let $M$ be a realisation system \ref{minus-three-points-1.9}.
We spell out what a realisation system $E$ given by an extension of $\QQ(0)$ by $M$ is, when $M$ is of weight $\leqslant-1$ (i.e. $M=W_{-1}(M)$).

@@

::: {.trmenv #M5P title="(M5)P" latex="{(M5)P}"}
The comparison morphisms for $E$ can be identified with comparison morphisms $P_\BB\otimes\CC\simto P_\DR\otimes\CC$, $P_\BB\otimes\AAf\simto P_\AAf$, and $P_\DR\otimes\QQ_p\simto P_\crisp$ that are compatible with the analogous morphisms for $M$.
*Notation:* $\otimes$, see \ref{minus-three-points-0.4}.
:::
